some title

### 2018-02-12 2018/chapters/4_distributions.tex
normal, t, f

@@

\begin{theorem}
Consider a random vector $z \in \mathbb{R}^n$ which components are independent
and follow standard normal distribution, $z_i \sim \mathcal{N}(0,1)$.
Consider also a fixed $k$-dimensional subspace $L$ in $\mathbb{R}^n$.
Let the projection of vector $z$ onto the subspace $L$ be $\hat z$ and
its length squared $Q$
\[
Q = \lVert \hat z \rVert^2 = \langle \hat z, \hat z \rangle = \hat z^T \hat z
\]
Then $Q$ follows the chi-squared distribution with $k$ degrees of freedom.
\end{theorem}

@@

Another nice property of~the~hat-mattix is symmetry.
Thus, it can be decomposed as
\[
H = P D P^T,
\]
where we choose the vectors of matrix $P$ to be unit and orthogonal,
and $D = diag{\lambda}$ where $\lambda_i$ is an eigenvalue of $H$.

@@

@@

>>>>

By contrast with substantial majority of books, the univariate normal distribution
can be derived form the multivariate normal distribution.
In this section we show how to obtain the univariate normal from the bivariate.

\begin{marginfigure}
  \includegraphics[width=\linewidth]{figures/04_normal.pdf}
  \caption{Black dots represent the gas molecules.
  The red dot stands for the one we catch.
  Its speed along the horizontal axis is $v_1$, i.e., the first component of
  the velocity vector, and its speed along the vertical axis is $v_2$.}
\end{marginfigure}

Assume there are gas molecules moving chaotically on a plane
and we can measure the velocity vector of one of them.
We denote this vector as $V = \begin{pmatrix} X \\ Y \end{pmatrix}$
where $X$ and $Y$ satnd for the horizontal and vertical components respectively.
Assume additionally that
\begin{enumerate}
  \item the joint distribution finction $f(x,y)$ does not depend on
  the vector $(X,Y)^T$ direction but depends on its length only,
  \item the orthogonal components of the velocity vector are independent.
\end{enumerate}

\marginnote{
It is more common to introduce a standard normal distribution in terms of its PDF
\begin{definition}
A continuous random variable $\xi$ has a standard normal distribution
if its PMF is given by
\[
f_{\xi} (x) = \frac{1}{\sqrt{2\pi}}\exp\left(-\frac{x^2}{2}\right).
\]
\end{definition}
After that multivariate normal is defined.
\begin{definition}
Let $\xi_i \stackrel{iid}{\sim} \mathcal{N}(0, 1)$ then $\xi \sim \mathcal{N}(\vec{0}, I)$
where $\xi = \begin{pmatrix} \xi_1 \\ \vdots \\ \xi_n \end{pmatrix}$,
$I$ is $n \times n$ identity matrix, and its PMF is
\[
f_{\xi} (x_1, \ldots, x_n) = \frac{1}{(\sqrt{2\pi})^n}\exp\left(-\frac{x_1^2+\ldots+x_n^2}{2}\right).
\]
\end{definition}
And finally, location-scale transformations are applied.
}

\begin{theorem}
The only way the assumptions (1) and (2) are satisfied is when
$X \sim \mathcal{N}(0, \sigma^2)$, $Y \sim \mathcal{N}(0, \sigma^2)$
and $X$, $Y$ are independent.
\end{theorem}

\begin{proof}
First of all, consider a vector $V' = \begin{pmatrix} -Y \\ X \end{pmatrix}$, i.e.,
the original vector $V$ rotated $90^{\circ}$ counterclock-wise.
By the assumption (1), this operation did not change the distribution of $V$.
Hence, $V' \sim V$ which implies $-Y \sim X$ and $X \sim Y$.
It follows that
\[
\begin{cases}
\E(-Y) = \E(X) \\
\E(X) = \E(Y)
\end{cases}
\]
which holds for $\E(X) = \E(Y) = 0$ only.

Likewise, $\Var(X) = \Var(Y)$ and here we additionally assume
without loss of generality that $\Var(X) = \Var(Y) = 1$.

Next, we introduce the angle between $V$ and the horizontal axis $U$ and
the length of the velocity vector $R = \sqrt{X^2 + Y^2}$.
Obviously, $X = R \cos U$ and $Y = R \sin U$.
Note that since the joint distribution of $X$ and $Y$ depends only on
the length of vector $V$ the distribution function of $U$ can only be constant,
thus $U \sim Unif(0, 2\pi)$.

Applying assumption (1) again, we conclude that the joint distribution function
can be written as a function of the length of the velocity vector, or equivalently
of the length squared:
\[
f(x,y) = h(x^2 + y^2).
\]
By the assumption (2), orthogonal components of $V$ are independent.
Hence, the joint PMF can be decomposed to the product of marginal ones:
\[
f(x,y) = f(x)f(y) = g(x^2)g(y^2)
\]
where the latter equality was written for convenience.
Putting everything together, we obtain
\[
h(x^2 + y^2) = g(x^2)g(y^2).
\]
Next, we take the derivative of both sides with respect to $y^2$ and
then substitute $y^2 = 0$ to get a constant $k$:
\begin{align*}
h'(x^2 + y^2) &= g(x^2)g'(y^2) \\
h'(x^2) &= g(x^2)g'(0) \\
h'(x^2) &= k \cdot g(x^2)
\end{align*}

\marginnote{
In order to obtain $k$ we computed
\begin{align*}
\E(X^2) &= \int_{-\infty}^{\infty} x^2 \sqrt{c} e^{kx^2} dx \\
&= \left. x \cdot \sqrt{c} e^{kx^2} \cdot \frac{k}{2} \right|_{-\infty}^{\infty} - \int_{-\infty}^{\infty} \sqrt{c} e^{kx^2} \cdot \frac{1}{2k} dx \\
&= - \frac{1}{2k} \int_{-\infty}^{\infty} \sqrt{c} e^{kx^2} \\
&= - \frac{1}{2k} \cdot 1 \\
&= 1.
\end{align*}
}

Solving the differential equation, we obtain
\[
h(x^2) = c e^{kx^2}, \quad c \in \mathbb{R}.
\]
So the joint PMF can be written as follows:
\[
f(x,y) = h(x^2 + y^2) = c e^{k(x^2+y^2}
\]
and due to independece of $X$ and $Y$ the PMF of $X$ is
\[
f(x) = \sqrt{c} e^{kx^2}.
\]

\marginnote{
In order to obtain $c$ we computed:
\begin{align*}
\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} e^{-\frac{x^2 + y^2}{2}} dx dy &= \int_{0}^{2\pi} \int_{0}^{\infty} e^{-\frac{r^2}{2}}r dr d\theta \\
&= \int_{0}^{2\pi} \left(\int_{0}^{\infty} e^{-u} du \right) d\theta \\
&= \int_{0}^{2\pi} 1 d\theta \\
&= 2 \pi.
\end{align*}
}

In oreder to find the constant $k$, we need to solve $\E(X^2) = 1$.
Computing the integral we obtain $k=-\frac{1}{2}$.

Finally, we need to normalize $f(x) = c e^{-\frac{x^2 + y^2}{2}}$ so as to
obtain $c$. Again, computing another integral,
we conclude that $c=(2\pi)^{-1}$ which finishes the proof.
\end{proof}

Notice, that any other $\mathcal{N}(\mu, \sigma^2)$ can be obtained by applying
location-scale transformations.

The theorem can be generalized to the n-dimensional case.
\begin{theorem}\label{th:mvn}
The vector $z = \begin{pmatrix} z_1 \\ \vdots \\ z_n \end{pmatrix}$
folllows the standard multivariate normal distribution and its components
are independent if and only if
\begin{enumerate}
  \item f(z) depends on $\vert z \vert$ only,
  \item the projections of vector $z$ onto the orthogonal subspaces $A$ and $B$
  in $\mathbb{R}^n$ are independent.
\end{enumerate}
\end{theorem}

@@

\begin{theorem}\label{th:chi}
Consider a random vector $z \in \mathbb{R}^n$ which components are independent
and follow standard normal distribution, $z_i \sim \mathcal{N}(0,1)$.
Consider also a fixed $k$-dimensional subspace $L$ in $\mathbb{R}^n$.
Let the projection of vector $z$ onto the subspace $L$ be $\hat z$ and
its length squared $Q$
\[
Q = \lVert \hat z \rVert^2 = \langle \hat z, \hat z \rangle = \hat z^T \hat z
\]
Then $Q$ follows the chi-squared distribution with $k$ degrees of freedom.
\end{theorem}

@@

Another nice property of~the~hat-mattix is symmetry.
Thus, it can be decomposed as
\[
H = P D P^T,
\]
where we choose the vectors of matrix $P$ to be unit and orthogonal,
and $D = diag{(\lambda_1, \ldots, \lambda_n)}$ where $\lambda_i$ is an eigenvalue of $H$.

@@

\marginnote{
A continuous random variable $T$ has Student's distribution with $k$ degrees
of freedom if it can be expressed as
\[
T = \frac{Z}{\gamma_k/k},
\]
where $Z \sim \mathcal{N}(0,1)$, $\gamma_k \sim \chi^2_{k}$ and
$Z$, $\gamma_k$ are independent.
}

\begin{definition}
Let  $z = \begin{pmatrix} z_0 \\ z_1 \\ \vdots \\ z_k \end{pmatrix}$
where $z_i \sim \mathcal{N}(0, \sigma^2), i=0, \ldots, k$ and are independent.
Let $L_1$ be 1-dimensional subspace in $\mathbb{R}^{k+1}$, $L_2$ an orthogonal to $L_1$ subspace.
Then
\[
T = \frac{\lVert H_1 z \rVert}{\lVert H_2 z \rVert / \sqrt{dim L_2}}
\]
where $\lVert H_1 z \rVert$ is the length of projection onto $L_1$ of the vector $z$,
$\lVert H_2 z \rVert$ — the length of the projection onto $L_2$,
follows Student's distribution with $dim L_2$ degrees of freedom.
\end{definition}

Let us choose two orthogonal subspaces: one-dimensional $L_1$ and
$k$-dimensional $L_2$.

Previously we showed that, the squared length of projection follows
the chi-squared distribution with the degrees of freedom equal to the dimension
onto which the vector was projected. Thus, $\lVert H_1 z \rVert^2 \sim \chi^2_1$
and $\lVert H_2 z \rVert^2 \sim \chi^2_{k}$.
Now we can express $T^2$ as a ratio of the per-dimension lengths squared:
\[
T^2 = \frac{\lVert H_1 z \rVert^2}{\lVert H_2 z \rVert^2 \ dim L_2}
\]
Taking the square root of both sides, we obtain:
\[
T = \frac{\lVert H_1 z \rVert}{\lVert H_2 z \rVert  \sqrt{dim L_2}} = \tg \varphi
\]
The latter equlity can be illustarted with a $3$-dimensional example (see Figure~\ref{fig:f_dist}).

@@

\marginnote{
Generally, the following definition is given.
\begin{definition}
Let $\gamma_1 \sim \chi^2_{k_1}$, $\gamma_2 \sim \chi^2_{k_2}$,
$\gamma_1$, $\gamma_2$ independent.
Then
\[
\frac{\gamma_1/k_1}{\gamma_2/k_2} \sim F_{k_1, k_2}.
\]
\end{definition}
}

\begin{definition}\label{def:f}
Let $z = \begin{pmatrix} z_1 \\ \vdots \\ z_n \end{pmatrix}$
where $z_i \sim \mathcal{N}(0, \sigma^2)$ and are independent.
Let $L_1$, $L_2$ be orthogonal subspaces in $\mathbb{R}^n$.
then
\[
F = \frac{\lVert H_1 z \rVert^2 / dim L_1}{\lVert H_2 z \rVert^2 / dim L_2} \sim F_{dim L_1, dim L_2},
\]
where $\lVert H_1 z \rVert^2$, $\lVert H_2 z \rVert^2$ are the squared lengths
of $z$ projected onto $L_1$ and $L_2$ respectively.
\end{definition}

\marginnote{
numerical example?
}

Recall that by Theorem~\ref{th:mvn} the projections of a standard noraml vector
onto orthogonal subspaces are independent.
Thus, in terms of the Definition~\ref{def:f} $H_1 z$ and $H_2 z$ are independent.
Next, from the Theorem~\ref{th:chi} where we defined the chi-squared distribution
it follows that the squared lengths of these projections follow
the chi-squared distribution with the number of degrees of freedom
equal to the dimension of the subspace onto which the vector was projected.
In other words, $\lVert H_1 z \rVert^2 \sim \chi^2_{dim L_1}$,
$\lVert H_2 z \rVert^2 \sim \chi^2_{dim L_2}$.

Taking the ratio of these length squared, we get the interpretaion
of the angle between the original vector $z$ and its projection onto $L_1$:
\[
tg^2 \varphi = \frac{\lVert H_1 z \rVert^2}{\lVert H_2 z \rVert^2}.
\]
Adjusting this ratio to the degrees of freedom, we get the desired definition.

\begin{figure}
\centering
\includegraphics[width=0.55\linewidth]{figures/04_f_dist_example.pdf}
\caption{F-distribution as the ratio of the projection lengths squared
adjusted to the dimensions of the subspaces.}
\label{fig:f_dist}
\end{figure}
